International Transport costs:\\New Findings from modeling additive costs

### 2018-09-12 redaction/Presentations/Seminaire_Kiel_Sept2018/slides_Kiel_sept_2018.tex
Présentation KIel
Slides actualisés avec derniers résultats pour la présentation à Kiel

\documentclass[10 pt,Helvetica, french]{beamer}
% option compress pour réduire la taille des bars
\mode<presentation>
%[compress,hideallsubsections,width=<0.5cm>]
%\usetheme{Darmstadt}
%\usetheme{Malmoe}
%\usetheme{Warsaw}
%\useoutertheme{split}
%\usecolortheme{whale}

@@

\date[October 2017]{BETA Seminar, University of Nancy, October 2017}
\title[Trade Costs Black Box]{Beyond the Iceberg Hypothesis: \\Opening the Black Box of Transport Costs}
\author[Daudin et al.]{Guillaume Daudin\\
{\footnotesize Universit\'{e} Paris-Dauphine, PSL \& OFCE }\\ \smallskip
J\'{e}rome H\'{e}ricourt \\
{\footnotesize Universit\'{e} de Lille I \& CEPII }\\  \smallskip
Lise Patureau \\
{\footnotesize  Universit\'{e} Paris-Dauphine, PSL}}

@@

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item Trade costs have a central role in international economic analysis \vspace{0.1cm}
\begin{itemize}
\item[-] Declining over the second half of the 20$^{th}$ century (Jacks et al., 2008, Novy, 2013) \vspace{0.1cm}
\item[-] But still significant: Average international trade costs = a 74\% markup over production costs (Anderson \& Van Wincoop, 2004)
\end{itemize}
\item What exactly are ``trade costs''?  \vspace{0.1cm}
\begin{itemize}
\item[-] Transaction costs, policy costs, time costs, and transport costs \textit{per se}
\end{itemize}
\item Transport costs are a sizeable share of international trade costs \vspace{0.1cm}
\begin{itemize}
\item[-]  Amount to $\simeq 30\%$ of international trade costs  $\Leftrightarrow$ A 21\% markup over production costs  \vspace{0.1cm}
%\item[-] Elasticity of trade wrt freight costs = -3 (Behar \& Venables, 2011) \vspace{0.1cm}
\end{itemize}
\item[$\Rightarrow$] If much trade policy barriers have been removed, the transport cost component of trade costs remains sizeable \vspace{0.1cm}
\end{itemize}
\textbf{The paper: On international transport costs}

@@

\begin{frame}
\frametitle{Motivation (cont')}
\begin{itemize}
\item Standard modeling of trade costs: As an ad-valorem tax-equivalent \vspace{0.1cm}
\begin{itemize}
\item[-] As a constant percentage of the producer price per unit traded \vspace{0.1cm}
\item[$\Leftrightarrow$] Part of the ``iceberg cost'' hypothesis (Samuelson, 1954) \vspace{0.1cm}
\begin{itemize}
\item[$\ast$] With $p$ the import price, $\widetilde{p}$ the export price, $q$ the quantity traded and $\tau >1$ the trade cost \vspace{0.1cm}
\item[$\Rightarrow$] Multiplicative costs:     \footnotesize $pq = \tau \widetilde{p}q$
\normalsize
\end{itemize}
\end{itemize}
\item Yet... A debated question \vspace{0.1cm}
\item Would not trade costs rather exhibit an additive structure ?  \vspace{0.1cm} \vspace{0.1cm}
\begin{itemize}
\item[-] Why would it be more costly to transport from Milan to Paris, a pair of Italian shoes at price \euro 300 than a pair of Italian shoes at price \euro 50? \vspace{0.1cm}
\item[-] Pricing shipping often includes an additive component \vspace{0.1cm}
\begin{itemize}
\item[$\ast$] UPS: A \$125 fee charged for a 2 pound package from Oslo to NYC (Irrarazabal et al., 2015) \vspace{0.1cm}
%\item[$\ast$] Some trade policy instruments as well (quota license price) \vspace{0.1cm}
\item[$\Rightarrow$] Additive costs: \footnotesize $pq = (t+\widetilde{p})q  $
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\begin{itemize}
\item Recent empirical evidence in support of the additive structure of trade costs \vspace{0.1cm}
\begin{itemize}
\item[-] Irrarazabal et al. (2015), Hummels \& Skiba (2004), Martin (2012)  \vspace{0.2cm}
\end{itemize}
\item The structure (additive vs iceberg) of transport costs is not anecdotal \vspace{0.1cm}
\begin{itemize}
\item[-] Additive costs play an important role in shaping the pattern of trade flows (Alchian \& Allen, 1964) \vspace{0.1cm}
\item[-] Strong normative implications, notably w.r.t. the welfare gains of trade liberalization (Sorensen, 2014)\vspace{0.1cm}

\end{itemize}
\item[$\Rightarrow$] Transport costs are likely to display an additive component, but precisely... by how much? \vspace{0.2cm}
\end{itemize}
\textbf{One objective of the paper: Provide an answer to this question}

@@

\begin{frame}
\frametitle{Our paper in one question (and 3 answers)}

\begin{itemize}
\item  \textbf{Do additive transport costs matter?} \vspace{0.1cm}
\begin{itemize}
\item[-] Provide an empirical decomposition of the structure of transport costs \vspace{0.1cm}
\item[-] Using the US imports database over 1974-2013 (air/vessel)\vspace{0.2cm}
\end{itemize}
\item[$\Rightarrow$] \textbf{Yes, they do} \vspace{0.1cm}

\item[(1)] \textbf{In terms of size}: Additive costs are sizeable \vspace{0.1cm}
\begin{itemize}
%\footnotesize{
\item[-] Additive cost: 1.8\% and 2.9\% of the export price, in air / vessel \vspace{0.1cm}
%\item[-] Iceberg cost: 2.5\% of the export price in air, and 3.2\% in vessel (mean value over 1974-2013) \vspace{0.1cm}
\item[-] Roughly 50\% of the overall transport costs \vspace{0.1cm}

%}
\end{itemize}
\item[(2)] \textbf{In terms of quality of fit}: With the additive component included, \vspace{0.1cm}
\begin{itemize}
%\footnotesize{
\item[-] The estimated iceberg component is reduced by a factor of 2 \vspace{0.1cm}
\item[-] A substantially better ``goodness-of-fit'' %}
%}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\begin{itemize}
\item[(3)] \textbf{In accounting for the time trend of transport costs} \vspace{0.2cm}
\begin{itemize}
\item[-] The modelling of the additive component of transport costs is key \vspace{0.1cm}
\item[-] Because it changes the result of excluding the trade composition effects when characterizing the time trend \vspace{0.1cm}
\begin{itemize}
\item[$\ast$] The decrease in TC over time is mostly attributable to the decrease in the ``pure'' transport costs,  \vspace{0.1cm}
\item[$\ast$] Rather than to trade composition effects  \vspace{0.1cm}
\end{itemize}
\item[-] A result in sharp contrast with Hummels (2007) \vspace{0.2cm}
\end{itemize}
\end{itemize}
\textbf{All our results: Provide new quantitative evidence about the importance of the additive component in international transport costs}

\begin{frame}
\frametitle{Related literature: Two related strands}
\begin{itemize}
\item Challenging the dominant role of iceberg costs in international trade \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Hummels and Skiba (2004) (sectoral data), Martin (2012) (firm-level data)
\item[-] The closest to us: Irarrazabal, Moxnes and Opromolla (2015) \vspace{0.1cm}
\item[$\Rightarrow$] In support of the additive cost component assumption \vspace{0.1cm}
\item[$\neq$] Our results: Quantitatively estimate the \textit{value} of the additive costs \vspace{0.1cm}
\item[-] A broader view of the magnitude of additive costs \vspace{0.1cm}
%\begin{itemize}
%\item[$\star$] US import flows over 1974-2013 \vspace{0.1cm}
%\end{itemize}
\item[$\Rightarrow$] Additive costs do matter quantitatively, throughout the period } \vspace{0.1cm}
\end{itemize}
\item Studying the patterns of trade costs over time \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Hummels (2007), Behar and Venables (2011) \vspace{0.1cm}
\item[-] What lies behind the decrease in transportation costs observed in the recent decades? \vspace{0.1cm}
\item[-] Hummels (2007): Mostly due to trade composition effects \vspace{0.1cm}
\item[$\neq$] Our results: Trade composition effects only a minor role \vspace{0.1cm}
\item[-] Why such a difference?  \vspace{0.1cm}
\item[$\Rightarrow$]Key role of the modelling of the additive component}
\end{itemize}

@@

\begin{frame}
\frametitle{Plan of the talk}
\begin{itemize}
\item Data Sources  \vspace{0.2cm}
\item Empirical Methodology \vspace{0.2cm}
\item Results \vspace{0.2cm}
\item Conclusion
\end{itemize}
\end{frame}

\begin{frame} [label=slide_data]
\frametitle{Data sources}
\begin{itemize}
\item Our measure of international transport costs: The difference between the ``export'' price and the ``import'' price of US imports\vspace{0.1cm}
\item Database: US Imports of Merchandise database \vspace{0.1cm}
\begin{itemize}
\item[-] The export (fas) price, $\widetilde{p}$: the price for one kg of merchandise at the country export point \vspace{0.1cm}
\item[-] The import (cif) price, $p$: the price for one kg of merchandise at the entry in the US \vspace{0.1cm}
\item[-] Yearly basis, from 1974 to 2013, SITC 5-digit classification level, by transport mode (air or vessel) \vspace{0.1cm}
%(Could go up to HS-10 with recent data)\vspace{0.1cm}
\end{itemize}
\item[$\Rightarrow$] Our dependent variable: Based on the ratio $p/\widetilde{p}$ \vspace{0.1cm}
\item The RHS is at the 3-digit classification level  \vspace{0.1cm}
\begin{itemize}
\item[-] Estimation at the 4-digit level on some selected years as robustness \vspace{0.1cm}
\item[-] Approximatively 200 sectors (3 digits), from around 200 countries  \vspace{0.1cm}
\begin{itemize}
\item[$\ast$] Around 600-700 sectors at the 4-digit level
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{More on our database}
\begin{itemize}
\item Implications (and limitations) \vspace{0.1cm}
\begin{itemize}
\item[-] Only cover international transport costs \vspace{0.1cm}
\item[-] Among transport costs, insurance + handling + quantitative freight costs (e.g. not those related to the time value of goods) \vspace{0.1cm}
%\item[$\Leftrightarrow$] Among the 21\% markup from AVW, $\simeq$ 11\% for freight costs (the remaining 9\% for time costs). But ``rough'' estimate. Precisely?  \vspace{0.1cm}
\end{itemize}
\item A rich database to exploit \vspace{0.1cm}
\begin{itemize}
\item[-] US imports, long time period: Broad view of international trade flows \vspace{0.1cm}
\item[-] A reliable database, already used by Hummels (2007), but on a shorter length of time \vspace{0.1cm}
\item[-] Have both the import and the export prices  \vspace{0.1cm}
\item[$\Rightarrow$] Allows the estimate the levels of both the ad-valorem and the additive transport costs  \vspace{0.1cm}
\begin{itemize}
\item[$\neq$] Irarrazabal et al. (2015)
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Empirical specification (1)}
\textbf{The equation at the root of the estimation}
\begin{itemize}
\item Relate the import price $p$ to the export price $\widetilde{p}$ given both additive (per-kg) costs $t$ and ad-valorem costs $\tau$:
$$p = \tau \widetilde{p} + t, \qquad \text{with}\quad \tau \geq 1,\quad t \geq 0$$
\item From country $i$, for a product $k$ (at the $k=$ 5-digit classification level, and for a given transport mode)   \vspace{0.1cm}
\item Rewrite to get:
\begin{equation}
\frac{p_{ik}}{\widetilde{p}_{ik}} -1 = \tau_{ik} -1 +\frac{t_{ik}}{ \widetilde{p}_{ik}} \label{eq:basis_equation}
\end{equation}
\item[$\Rightarrow$] For each year over 1974-2013\vspace{0.1cm}
\begin{itemize}
\item[-] The equation is also mode (air or vessel) - specific
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}[label = slide_variancedecomp]
\frametitle{Empirical specification (2)}
\textbf{The estimation strategy} \vspace{0.1cm}
\begin{itemize}
\item Make some simplifying assumptions to get the estimated equation \vspace{0.1cm}
\item About the specification of both additive and multiplicative costs  \vspace{0.1cm}
\begin{itemize}
\item[(1)] Separability between the origin country $i$ and the product $k$ dimensions (as in Irarrazabal et al., 2015)\vspace{0.1cm}
\item[(2)] All products $k$ in a 3-digit sector $s$ share the same structure of costs \hyperlink{app_variancedecomp}{\beamergotobutton{More on this}}\vspace{0.1cm}
\item[$\Leftrightarrow$] Write $t_{ik}$ and $\tau_{ik}$ as:
\footnotesize
\begin{equation}
 \tau_{ik} = \tau_{i} \times \tau_{s(k)}, \qquad t_{ik} = t_{i} + t_{s(k)} \label{eq:specifTC}
 \end{equation}
 \normalsize
\end{itemize}
\item About the specification of the error term \vspace{0.1cm}
\begin{itemize}
\item[-] Given the constraint $\frac{p_{ik}}{\widetilde{p}_{ik}} -1>0$, the error term should be always positive and multiplicative \vspace{0.1cm}
\item[$\Rightarrow$] The equation of interest becomes:
\footnotesize
$$\frac{p_{ik}}{\widetilde{p}_{ik}}-1 =\left(\tau_{i} \times \tau_{s(k)} -1+\frac{t_{i} + t_{s(k)}}{\widetilde{p}_{ik}} \right)\times \exp(\epsilon_{ik})$$
\normalsize
\begin{itemize}
\item[-] With $\epsilon_{ik}$ following a normal law centered on 0
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

\begin{frame}
\begin{itemize}
\item Taking logs, we finally estimate the following equation
\footnotesize
\begin{equation}
\ln\left(\frac{p_{ik}}{\widetilde{p}_{ik}}-1 \right)= \ln \left(\tau_{i} \times \tau_{s(k)}+\frac{t_{i} + t_{s(k)}}{\widetilde{p}_{ik}}-1 \right) + \epsilon_{ik} \label{eq:est_equation}
\end{equation}
\normalsize
\begin{itemize}
\item[-] $\tau_i$, $\tau_{s(k)}$, $t_i$, $t_{s(k)}$ are the parameters (i.e., fixed effects) to be estimated \vspace{0.1cm}
\end{itemize}
\item A non-linear equation (due to the additive costs)  \vspace{0.1cm}
\item[$\Rightarrow$] Estimation using non-linear squares \vspace{0.1cm}
\begin{itemize}
\item[-] At the basis of the method: Approximate the model by a linear one and refine the parameters by successive iterations \vspace{0.1cm}
\item[-] The criterion for convergence: That the sum of the squares of the residuals does not increase from one iteration to the next \vspace{0.1cm}
\end{itemize}
\item Eliminate potential influence of outliers: Exclude the 5 percent of the upper and lower tails of the distribution
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Empirical specification (3)}
\textbf{How to characterize the importance of additive costs relatively to ad-valorem?} \vspace{0.1cm}
\begin{itemize}
\item Estimate Equation (\ref{eq:est_equation}) constraining $t=0$ \vspace{0.1cm}
\end{itemize}
$\Rightarrow$ Estimate \textit{two} models\vspace{0.1cm}
\begin{itemize}
\item[(A)] With additive costs excluded (only ad-valorem costs):
\footnotesize
\begin{equation}
\ln\left(\frac{p_{ik}}{\widetilde{p}_{ik}}-1 \right)= \ln \left(\tau_{i} \times \tau_{s(k)}-1 \right) + \epsilon^{ice}_{ik} \label{eq:model_nlI}
\end{equation}
\normalsize
\item[(B)] With additive costs included:
\footnotesize
\begin{equation}
\ln\left(\frac{p_{ik}}{\widetilde{p}_{ik}}-1 \right)= \ln \left(\tau_{i} \times \tau_{s(k)}+\frac{t_{i} + t_{s(k)}}{\widetilde{p}_{ik}}-1 \right) + \epsilon_{ik} \label{eq:est_equation}
\end{equation}
\normalsize
\begin{itemize}
\item[$\ast$] Under Model (B), $\simeq$ 800 fixed effects to estimate  \vspace{0.1cm}
\item[$\ast$] For each year over 1974-2013, by transport mode (air/vessel)
%\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

>>>>

\documentclass[10 pt,Helvetica, french]{beamer}
%\documentclass[10 pt,Helvetica, french, handout]{beamer}
% option compress pour réduire la taille des bars
\mode<presentation>
%[compress,hideallsubsections,width=<0.5cm>]
%\usetheme{Darmstadt}
%\usetheme{Malmoe}
%\usetheme{Warsaw}
%\useoutertheme{split}
%\usecolortheme{whale}

@@

\date[September 2018]{Lunch seminar in Interational Economics, Kiel \\ September 7, 2018}
\title[New Findings from modeling additive costs]{International Transport costs:\\New Findings from modeling additive costs}
\author[Daudin et al.]{Guillaume Daudin\\
{\footnotesize Universit\'{e} Paris-Dauphine, PSL \& OFCE }\\ \smallskip
J\'{e}rome H\'{e}ricourt \\
{\footnotesize Universit\'{e} de Lille \& CEPII }\\  \smallskip
Lise Patureau \\
{\footnotesize  Universit\'{e} Paris-Dauphine, PSL}}

@@

\begin{frame}
\frametitle{Motivation}
\begin{itemize}
\item Trade costs have a central role in international economic analysis \vspace{0.1cm}
\begin{itemize}
\item[-] Declining over the second half of the 20$^{th}$ century (Jacks et al., 2008, Novy, 2013) \vspace{0.1cm}
\item[-] But still significant: Average international trade costs = a 74\% markup over production costs (Anderson \& Van Wincoop, 2004)

\pause
\end{itemize}
\item What exactly are ``trade costs''?  \vspace{0.1cm}
\begin{itemize}
\item[-] Transaction costs, policy costs, time costs, and... transport costs \textit{per se}
\end{itemize}
\pause
\item Transport costs are a sizeable share of international trade costs \vspace{0.1cm}
\begin{itemize}
\item[-]  Amount to $\simeq 30\%$ of international trade costs  $\Leftrightarrow$ A 21\% markup over production costs  \vspace{0.1cm}
%\item[-] Elasticity of trade wrt freight costs = -3 (Behar \& Venables, 2011) \vspace{0.1cm}
\end{itemize}
\item[$\Rightarrow$] If much trade policy barriers have been removed, the transport cost component of trade costs remains sizeable \vspace{0.1cm}
\end{itemize}
\textbf{This paper: Focus on international transport costs}

\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Motivation (cont')}
\begin{itemize}
\item Standard modeling of trade costs: As an ad-valorem tax-equivalent \vspace{0.1cm}
\begin{itemize}
\item[-] As a constant percentage of the producer price per unit traded \vspace{0.1cm}
\item[$\Leftrightarrow$] Part of the ``iceberg cost'' hypothesis (Samuelson, 1954) \vspace{0.1cm}
\begin{itemize}
\item[$\ast$] With $p$ the import price, $\widetilde{p}$ the export price, $q$ the quantity traded and $\tau >1$ the trade cost \vspace{0.1cm}
\item[$\Rightarrow$] \textbf{Multiplicative costs: }    \footnotesize $\mathbf{pq = \tau \widetilde{p}q}$
\normalsize
\end{itemize}
\end{itemize}
\pause

\item Yet... A debated question \vspace{0.1cm}%\bigskip
\item Would not trade costs rather exhibit an additive structure ?  \vspace{0.1cm} 
\pause
\begin{itemize}
\item[-] Why would it be more costly to transport from Milan to Paris, a pair of Italian shoes at price \euro 300 than a pair of Italian shoes at price \euro 50? \vspace{0.1cm}
\item[-] Pricing shipping often includes an additive component \vspace{0.1cm}
\pause
\begin{itemize}
\item[$\ast$] UPS: A \$125 fee charged for a 2 pound package from Oslo to NYC (Irrarazabal et al., 2015) \vspace{0.1cm}
%\item[$\ast$] Some trade policy instruments as well (quota license price) \vspace{0.1cm}
\item[$\Rightarrow$] \textbf{Additive costs}: \footnotesize $\mathbf{pq = (t+\widetilde{p})q}  $
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\begin{itemize}
\item Recent empirical evidence in support of the additive structure of trade costs \vspace{0.1cm}
\begin{itemize}
\item[-] Irrarazabal et al. (2015), Hummels \& Skiba (2004), Martin (2012)  \vspace{0.2cm}
\end{itemize}
\pause
\item The structure (additive vs iceberg) of transport costs is not anecdotal \vspace{0.1cm}
\begin{itemize}
\item[-] Additive costs play an important role in shaping the pattern of trade flows (Alchian \& Allen, 1964) \vspace{0.1cm}
\item[-] Strong normative implications, notably w.r.t. the welfare gains of trade liberalization (Sorensen, 2014)\vspace{0.1cm}
\pause
\end{itemize}
\item[$\Rightarrow$] Transport costs are likely to display an additive component, but precisely... by how much? \vspace{0.2cm}
\item[$\Rightarrow$] How does this affect our perception of the time dynamics of transport cost?
\end{itemize}
\textbf{Objectives of the paper: answer both questions}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{frame}
\frametitle{Our paper in a nutshell}

\begin{itemize}
\item  \textbf{What we do} \vspace{0.1cm}
\begin{itemize}
\item[-] Provide an empirical decomposition of the structure of transport costs \vspace{0.1cm}
\item[-] Using the US imports database over 1974-2013 (air/vessel)\vspace{0.2cm}
\end{itemize}
\item[$\Rightarrow$] \textbf{Question 1: do additive costs matter?} \vspace{0.1cm}

\pause

\item[(1)] \textbf{YES, THEY DO!}: Additive costs are sizeable \vspace{0.1cm}
\begin{itemize}
%\footnotesize{
\item[-] Additive cost: 1.8\% and 2.9\% of the export price, in air / vessel \vspace{0.1cm}
%\item[-] Iceberg cost: 2.5\% of the export price in air, and 3.2\% in vessel (mean value over 1974-2013) \vspace{0.1cm}
\item[-] Roughly 50\% of the overall transport costs \vspace{0.1cm}
\item[-] A substantially better ``goodness-of-fit'' when the additive component included
%}
\end{itemize}
%\item[(2)] \textbf{In terms of quality of fit}: With the additive component included, \vspace{0.1cm}
%\begin{itemize}
%\footnotesize{
%\item[-] The estimated iceberg component is reduced by a factor of 2 \vspace{0.1cm}
%\item[-] A substantially better ``goodness-of-fit'' %}
%}
%\end{itemize}
\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\begin{itemize}
\item[$\Rightarrow$] \textbf{Question 2: does this change our accounting for the time trend of transport costs?} \vspace{0.1cm}
\pause
\item[(2)] \textbf{YES, IT DOES!} \vspace{0.2cm}
\begin{itemize}
\item[-] The modelling of the additive component of transport costs is key \vspace{0.1cm}
\item[-] Because it changes the result of excluding the trade composition effects when characterizing the time trend \vspace{0.1cm}
\pause
\begin{itemize}
\item[$\ast$] Hummels (2007): composition effects partly offset the decrease in transport costs  \vspace{0.1cm} %XXX CAREFUL to misinterpretations here!!!XXX
\item[$\ast$] DHP (2018): composition effects do not matter much; if anything, they amplify (slightly) the decrease in pure transport costs.  \vspace{0.1cm}
\end{itemize}
%\item[-] A result in sharp contrast with Hummels (2007) \vspace{0.2cm}
\end{itemize}
\pause
\end{itemize}
\textbf{All our results: Provide new quantitative evidence about the importance of the additive component in international transport costs}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Related literature: Two related strands}
\begin{itemize}
\item Challenging the dominant role of iceberg costs in international trade \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Hummels and Skiba (2004) (sectoral data), Martin (2012) (firm-level data)
\item[-] The closest to us: Irarrazabal, Moxnes and Opromolla (2015) \vspace{0.1cm}
\item[$\Rightarrow$] In support of the additive cost component assumption \vspace{0.1cm}
\pause
\item[$\neq$] \textbf{Our results: Quantitatively estimate the \textit{value} of the additive costs} \vspace{0.1cm}
\item[-] A broader view of the magnitude of additive costs \vspace{0.1cm}
%\begin{itemize}
%\item[$\star$] US import flows over 1974-2013 \vspace{0.1cm}
%\end{itemize}

\item[$\Rightarrow$] Additive costs do matter quantitatively, throughout the period} \vspace{0.1cm}
\pause
\end{itemize}
\item Studying the patterns of trade costs over time \vspace{0.1cm}
\begin{itemize}
\footnotesize{
\item[-] Hummels (2007), Behar and Venables (2011) \vspace{0.1cm}
\item[-] What is exactly behind the decrease in transportation costs observed in the recent decades? \vspace{0.1cm}
\item[-] Hummels (2007): trade composition effects partly cancelled decrease in transport costs \vspace{0.1cm}
\pause
\item[$\neq$] \textbf{Our results: Trade composition effects only had a (very) minor impact on the trend} \vspace{0.1cm}
\item[-] Why such a difference?  \vspace{0.1cm}
\item[$\Rightarrow$]Key role of the modelling of the additive component}
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Roadmap}
\begin{itemize}
\item Data Sources  \vspace{0.2cm}
\item Empirical Methodology \vspace{0.2cm}
\item Results \vspace{0.2cm}
\item A few Robustness Checks \vspace{0.2cm}
\item Conclusion
\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%ù

\begin{frame} [label=slide_data]
\frametitle{Data sources}
\begin{itemize}
\item Our measure of international transport costs: The difference between the ``export'' price and the ``import'' price of US imports\vspace{0.1cm}
\item Database: US Imports of Merchandise database \vspace{0.1cm}
\pause
\begin{itemize}
\item[-] The export (fas) price, $\widetilde{p}$: the price for one kg of merchandise at the country export point \vspace{0.1cm}
\item[-] The import (cif) price, $p$: the price for one kg of merchandise at the entry in the US \vspace{0.1cm}
\item[-] Yearly basis, from 1974 to 2013, SITC 5-digit classification level, by transport mode (air or vessel) \vspace{0.1cm}
%(Could go up to HS-10 with recent data)\vspace{0.1cm}

\pause
\end{itemize}
\item[$\Rightarrow$] \textbf{Our dependent variable: Based on the ratio} $\mathbf{p/\widetilde{p}}$ \vspace{0.1cm}

\pause

\item The RHS is at the 3-digit classification level  \vspace{0.1cm}
\begin{itemize}
\item[-] Estimation at the 4-digit level on some selected years as robustness \vspace{0.1cm}
\item[-] Approximatively 200 sectors (3 digits), from around 200 countries  \vspace{0.1cm}
\begin{itemize}
\item[$\ast$] Around 600-700 sectors at the 4-digit level
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{More on our database}
\begin{itemize}
\item Implications (and limitations) \vspace{0.1cm}
\begin{itemize}
\item[-] Only cover international transport costs \vspace{0.1cm}
\item[-] Among transport costs, insurance + handling + quantitative freight costs (e.g. not those related to the time value of goods) \vspace{0.1cm}
%\item[$\Leftrightarrow$] Among the 21\% markup from AVW, $\simeq$ 11\% for freight costs (the remaining 9\% for time costs). But ``rough'' estimate. Precisely?  \vspace{0.1cm}
\pause
\end{itemize}
\item A rich database to exploit \vspace{0.1cm}
\begin{itemize}
\item[-] US imports, long time period: Broad view of international trade flows \vspace{0.1cm}
\item[-] A reliable database, already used by Hummels (2007), but on a shorter length of time \vspace{0.1cm}
\item[-] Have both the import and the export prices  \vspace{0.1cm}
\pause

\item[$\Rightarrow$] \textbf{Allows to estimate the levels of both the ad-valorem and the additive transport costs} \vspace{0.1cm}
\begin{itemize}
\item[$\neq$] Irarrazabal et al. (2015)
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\frametitle{Empirical specification (1)}
\textbf{The equation underlying the estimation}
\begin{itemize}
\item Relate the import price $p$ to the export price $\widetilde{p}$ given both additive (per-kg) costs $t$ and ad-valorem costs $\tau$:
$$p = \tau \widetilde{p} + t, \qquad \text{with}\quad \tau \geq 1,\quad t \geq 0$$
\item From country $i$, for a product $k$ (at the $k=$ 5-digit classification level, and for a given transport mode)   \vspace{0.1cm}
\item Rewrite to get:
\begin{equation}
\frac{p_{ik}}{\widetilde{p}_{ik}} -1 = \tau_{ik} -1 +\frac{t_{ik}}{ \widetilde{p}_{ik}} \label{eq:basis_equation}
\end{equation}
\item[$\Rightarrow$] For each year over 1974-2013\vspace{0.1cm}
\begin{itemize}
\item[-] The equation is also mode (air or vessel) - specific
\end{itemize}
\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}[label = est_strategy]
\frametitle{Empirical specification (2)}
\textbf{The estimation strategy} \vspace{0.1cm}
\begin{itemize}
\item Make some simplifying assumptions to get the estimated equation \vspace{0.1cm}
\item About the specification of both additive and multiplicative costs  \vspace{0.1cm}
\begin{itemize}
\item[(1)] Separability between the origin country $i$ and the product $k$ dimensions (as in Irarrazabal et al., 2015)\vspace{0.1cm}
\item[(2)] All products $k$ in a 3-digit sector $s$ share the same structure of costs \hyperlink{app_variancedecomp}{\beamergotobutton{More on this}}\vspace{0.1cm}
\item[$\Leftrightarrow$] Write $t_{ik}$ and $\tau_{ik}$ as:
\footnotesize
\begin{equation}
 \tau_{ik} = \tau_{i} \times \tau_{s(k)}, \qquad t_{ik} = t_{i} + t_{s(k)} \label{eq:specifTC}
 \end{equation}
 \normalsize
\end{itemize}

\pause

\item About the specification of the error term \vspace{0.1cm}
\begin{itemize}
\item[-] Given the constraint $\frac{p_{ik}}{\widetilde{p}_{ik}} -1>0$, the error term should be always positive and multiplicative \vspace{0.1cm}
\item[$\Rightarrow$] The equation of interest becomes:
\footnotesize
$$\frac{p_{ik}}{\widetilde{p}_{ik}}-1 =\left(\tau_{i} \times \tau_{s(k)} -1+\frac{t_{i} + t_{s(k)}}{\widetilde{p}_{ik}} \right)\times \exp(\epsilon_{ik})$$
\normalsize
\begin{itemize}
\item[-] With $\epsilon_{ik}$ following a normal law centered on 0
\end{itemize}
\end{itemize}
\end{itemize}
\end{frame}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{frame}
\begin{itemize}
\item Taking logs, we finally estimate the following equation
\footnotesize
\begin{equation}
\ln\left(\frac{p_{ik}}{\widetilde{p}_{ik}}-1 \right)= \ln \left(\tau_{i} \times \tau_{s(k)}+\frac{t_{i} + t_{s(k)}}{\widetilde{p}_{ik}}-1 \right) + \epsilon_{ik} \label{eq:est_equation}
\end{equation}
\normalsize
\begin{itemize}
\item[-] $\tau_i$, $\tau_{s(k)}$, $t_i$, $t_{s(k)}$ are the parameters (i.e., fixed effects) to be estimated \vspace{0.1cm}
\pause
\end{itemize}
\item A non-linear equation (due to the additive costs)  \vspace{0.1cm}
\item[$\Rightarrow$] Estimation using non-linear squares \vspace{0.1cm}
%\begin{itemize}
%\item[-] At the basis of the method: Approximate the model by a linear one and refine the parameters by successive iterations \vspace{0.1cm}
%%\item[-] The criterion for convergence: That the sum of the squares of the residuals does not increase from one iteration to the next \vspace{0.1cm}
%\end{itemize}
\pause
\item Eliminate potential influence of outliers: Exclude the 5 percent of the upper and lower tails of the distribution
\end{itemize}
\end{frame}
